---
title: <center> <h2>Tutorial of the MetaGSCA package</h2> </center>
date: <center> <h4>`r Sys.Date()` </center>
output: 
  html_document:
    fontsize: 20pt
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cerulean
---

<style type="text/css">

body {
  margin-left:430px;
  margin-right:100px;
}

#TOC {
  position: fixed;
  left: 0;
  top: 10px;
  bottom: 10px;
  width: 480px;
  height: 100%;
  overflow:auto;
  collapse: true;
}

</style>
  
<div style="margin-bottom:50px;">
</div>

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
library(knitr)
```

<font size="4">**Package:**</font>   
MetaGSCA 0.99.0  

<font size="4">**Authors:**</font>       
Hui Yu <HuiYu1@salud.unm.edu>  
Haocan Song <haocan.song@vumc.org>  
Yan Guo <YaGuo@salud.unm.edu>  
Fei Ye <fei.ye@vumc.org>   




# Introduction  

<font size="3"> This tutorial provides an overview and a tutorial of the R package *MetaGSCA* for Meta Gene Set Co-Expression Analysis. </font>    

<font size="3"> **MetaGSCA (Meta Gene Set Co-expression Analysis)** is a computational tool to systematically assess the co-expression disturbance of a specified gene set with rigorous meta-analysis justification. A nonparametric approach that accounts for the correlation structure between genes is used to test whether the gene set is differentially co-expresssed between the two conditions, and a **permutation test p-value** is computed. **Bootstrapping** is used to construct confidence intervals of point estimates. A **meta-analysis of single proportions** is then performed with two options: random-intercept logistic regression model and the inverse variance method, to yield conclusive results over a number of individual studies. The meta-analysis empowered gene set differential co-expression tool MetaGSCA is implemented in the form of both a web application and an R package. </font>      

<font size="3"> Packages **data.table, meta(metafor), grid, sjstats, igraph** are required. </font>  

<font size="3"> The analysis will start by loading package MetaGSCA. </font>       
```{r eval=FALSE}
library(MetaGSCA)
```

<br>





# How to use the MetaGSCA package

<font size="3"> This Section illustrates the typical procedure for applying the methods available in package **MetaGSCA**. </font>  
<br>



## Set Working Directory
<font size="3"> Create a temporary directory "workdir" under the user's current working directory. All output generated in this tutorial will be saved in "workdir".  </font>  

```{r eval=FALSE}
## set your workdir
workdir <- setwd()
```
<br>



## Example Data

<font size="3"> Load the example data:    
**1. lists of genesets (29 pathways)**  
**2. lists of datasets (KIRP & LIHC)**  
The 2 source of data are necessary for running the examples and case studies in this tutorial. </font>     

```{r echo=FALSE, include=FALSE}
library(MetaGSCA)
```

```{r warning =FALSE}
###########################
## step 1, read in pathways
mygeneset <- list.files(system.file("extdata/pathway29", package = "MetaGSCA"),
                        pattern = "*.lst", full.names=TRUE)
genesets <- list(); genesetnames <- list()
for (i in 1:length(mygeneset)) {
    g <- fread(mygeneset[i], header = FALSE, data.table = FALSE)
    genesets[[i]] <- as.character(g[, 1])
    genesetnames[[i]] <- tools::file_path_sans_ext(mygeneset[i])
}
genesetnames <- gsub(paste(system.file("extdata/pathway29", package = "MetaGSCA"),
                     "/", sep = ""), "", genesetnames)

##########################
## step 2,read in datasets
mydataset <- c(paste(system.file("extdata", package = "MetaGSCA"),
                     "/KIRP.rds", sep=""),
               paste(system.file("extdata", package = "MetaGSCA"),
                     "/LIHC.rds", sep=""))
datasets <- list(); groups <- list(); datasetnames <- list()
for (j in 1:length(mydataset)) {
    d <- readRDS(file = mydataset[j])
    datasets[[j]] <- d[-1,]
    groups[[j]] <- as.character(d[1, -1])
    datasetnames[[j]] <- tools::file_path_sans_ext(mydataset[j])
}
datasetnames <- gsub(paste(system.file("extdata", package = "MetaGSCA"),
                     "/", sep = ""), "", datasetnames)
```

<font size="3"> **Note:**   
**1. lists of genesets (29 pathways):**    
```{r echo=FALSE}
genesetnames
```

**2. lists of datasets (KIRP & LIHC):**   
*  **KIRP** (3819 rows & 65 columns) contain 3818 genes & 32 individuals, individuals are equally distributed in 2 groups (group1 & group2).    
*  **LIHC** (3687 rows & 101 columns) contain 3686 genes & 50 individuals, individuals are equally distributed in 2 groups (group1 & group2).    
</font>   
<br>



## Case Study
### Example 1: MetaGSCA 
#### Code

<font size="3"> **MetaGSCA** is used to perform meta & gene set co-expression analysis of one geneset.  </font>   

<font size="4"> **Arguments** </font>  
<font size="3"> 
*  ***list.geneset:*** **a** pre-specified gene list from a gene set or pathway.  
*  ***list.dataset:*** a list of datasets, first column is gene name.  
*  ***list.group:*** a list of samples/patients subgroup or condition (e.g. (1,1,1,2,2,2)).  
*  ***name.geneset:*** the name of the gene set, used for output file name.  
*  ***name.dataset:*** a list of dataset names corresponding to list.dataset, used for forest plot.  
*  ***nperm:*** number of permutations used to estimate the null distribution of the test statistic. If not given, a default value 500 is used.  
*  ***nboot:*** number of bootstraps used to estimate the point and interval estimate. If not given, a default value 200 is used.  
*  ***method.Inverse:*** logical. Indicating whether "Inverse" method is to be used for pooling of studies.  
*  ***method.GLMM:*** logical. Indicating whether "GLMM" method is to be used for pooling of studies.  
*  ***fixed.effect:*** logical. Indicating whether a fixed effect meta-analysis should be conducted.  
*  ***random.effect:*** logical. Indicating whether a random effects meta-analysis should be conducted.   
</font>     

```{r warning=FALSE, eval=FALSE}
## step 3.1,call MetaGSCA function to perform Meta-analysis
# dir.create(file.path(workdir, "result_single_geneset"), showWarnings = FALSE)
# setwd(file.path(workdir, "result_single_geneset"))
MetaGSCA(list.geneset = genesets[[2]],
         list.dataset = datasets,
         list.group = groups,
         name.geneset = genesetnames[[2]],
         name.dataset = datasetnames,
         nperm = 100,
         nboot = 100,
         method.Inverse = FALSE,
         method.GLMM = TRUE,
         fixed.effect = FALSE,
         random.effect = TRUE)
```
<br>

#### Result
```{r}
genesetnames[2]
genesets[2]
```
<br>

<font size="3"> There'll be 2 output file after **MetaGSCA**. 1 forest plot (.pdf) and 1 output data information file (.csv). </font>   
<br>
```{r echo=FALSE, out.width = "60%", fig.align='left', dpi=200}
result <- paste(system.file("extdata", package = "MetaGSCA"), "/single.png", sep="")
knitr::include_graphics(result)
```
<br>

<font size="3"> In the first file, we're able to get the forest plot of 2 datasets (KIRP & LIHC), with random effects model result. Attached with bootstrap result at the bottom side. </font> 
```{r echo=FALSE, out.width = "90%", fig.align='center', dpi=200, fig.cap="Gene[2]: Aminosugars metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot"}
result1 <- paste(system.file("extdata", package = "MetaGSCA"), "/Aminosugars metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot.png", sep="")
knitr::include_graphics(result1)
```
<br>

<font size="3"> In the second file, it includes the output data information of 2 datasets (KIRP & LIHC). </font> 
<center> Gene[2]: Aminosugars metabolism_Individual Dataset GSAR Results with Bootstrapping </center>  
```{r echo=FALSE}
result2 <- paste(system.file("extdata", package = "MetaGSCA"), "/Aminosugars metabolism_Individual Dataset GSAR Results with Bootstrapping.csv", sep="")
t(read.csv(result2))
```
<br>



### Example 2: MetaGSCA_Multi_Geneset

#### Code
<font size="3"> **MetaGSCA_Multi_Geneset** is used to perform meta & gene set co-expression analysis of multiple geneset.  </font>    

<font size="4"> **Arguments** </font>  
<font size="3"> 
*  ***list.geneset:*** **several** pre-specified gene lists from a gene set or pathway.    
*  ***name.geneset:*** the names of the gene sets, used for output file name.  
</font>  

```{r warning=FALSE, eval=FALSE}
## step 3.2,call MetaGSCA_Multi_Geneset function to perform Multiple geneset Meta-analysis
# dir.create(file.path(workdir, "result_multiple_geneset"), showWarnings = FALSE)
# setwd(file.path(workdir, "result_multiple_geneset"))
MetaGSCA_Multi_Geneset(list.geneset = genesets[2:3],
                       list.dataset = datasets,
                       list.group = groups,
                       name.geneset = genesetnames[2:3],
                       name.dataset = datasetnames,
                       nperm = 100,
                       nboot = 100,
                       method.Inverse = FALSE,
                       method.GLMM = TRUE,
                       fixed.effect = FALSE,
                       random.effect = TRUE)
```
<br>


#### Result

```{r}
genesetnames[2:3]
genesets[2:3]
```
<br>

<font size="3"> There'll be 4 output file after **MetaGSCA**. 2 forest plot (.pdf) and 2 output data information file (.csv). </font>   
<br>
```{r echo=FALSE, out.width = "60%", fig.align='left', dpi=200}
result <- paste(system.file("extdata", package = "MetaGSCA"), "/multiple.png", sep="")
knitr::include_graphics(result)
```
<br>


<font size="3"> In the 2 forest plot files, we're able to get the forest plot of 2 datasets (KIRP & LIHC), with random effects model result. Attached with bootstrap result at the bottom side. </font> 
```{r echo=FALSE, out.width = "90%", fig.align='center', dpi=200, fig.cap="Gene[2]: Aminosugars metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot"}
result1 <- paste(system.file("extdata", package = "MetaGSCA"), "/Aminosugars metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot.png", sep="")
knitr::include_graphics(result1)
```
<br>
```{r echo=FALSE, out.width = "90%", fig.align='center', dpi=200, fig.cap="Gene[3]: Arginine and Proline metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot"}
result1 <- paste(system.file("extdata", package = "MetaGSCA"), "/Arginine and Proline metabolism_GLMM_Meta Analysis for bootstrap p-value_Forest plot.png", sep="")
knitr::include_graphics(result1)
```
<br>


<font size="3"> In the 2 output data information files, they includes the output data information of 2 datasets (KIRP & LIHC). </font>   
<center> Gene[2]: Aminosugars metabolism_Individual Dataset GSAR Results with Bootstrapping </center>   
```{r echo=FALSE}
result2 <- paste(system.file("extdata", package = "MetaGSCA"), "/Aminosugars metabolism_Individual Dataset GSAR Results with Bootstrapping.csv", sep="")
t(read.csv(result2))
```
<br>
<center> Gene[3]: Arginine and Proline metabolism_Individual Dataset GSAR Results with Bootstrapping </center>
```{r echo=FALSE}
result2 <- paste(system.file("extdata", package = "MetaGSCA"), "/Arginine and Proline metabolism_Individual Dataset GSAR Results with Bootstrapping.csv", sep="")
t(read.csv(result2))
```
<br>


### Warning Result given by 2.3.1 & 2.3.2
<font size="3"> There might be a warning message if we perform **GLMM method** (method.GLMM = TRUE). If the oringinal p-values have the same value, GLMM method (function:metaprop) will automatically stop with an ERROR message, and bootstrap procedure couldn't be done. To deal with this problem, we set the oringinal p-value as the mean of different datasets p-value, with bootstrap results. </font>  
```{r echo=FALSE, out.width = "100%", fig.align='center', dpi=200}
warn <- paste(system.file("extdata", package = "MetaGSCA"), "/GLMM warning.png", sep="")
knitr::include_graphics(warn)
```
<br>
<br>


### Example 3: Pathway Crosstalk Analysis
For pathway crosstalk analysis, one call *PWcTalk* function in one line, or alternatively, execute several codelines to enable greater modulation of network layout. One must execute *MetaGSCA_Multi_Geneset* function on sufficiently many gene sets (say >100) before attempting pathway crosstalk analysis. To enable swift testing of pathway cross talk analysis, we provide one example input as if it was generated from the prior workflow.

```{r warning=FALSE, eval=FALSE}
data(input2PWcTalk)
## step 4.1, one code line to execute pathway crosstalk analysis. Simple, but no flexibility for layout tuning. 
PWcTalk(input2PWcTalk,test='binary',
  pTh.dataset=0.01,pTh.pwPair=0.01,pTh.pw=0.01,figname='PWcTalk',
  pdfW=10,pdfH=10,asp=0.7,vbase=15,ebase=2,vlbase=1,power=1/2)
## step 4.2, one code block to execute pathway crosstalk analysis. More code lines, but enabling interactive layout tuning.
preNW <- PWcTalkNWpre(input2PWcTalk,test='binary',
  pTh.dataset=0.01,pTh.pwPair=0.01,pTh.pw=0.01)
g_tkid <- PWcTalkNW(preNW$PW.pair,preNW$PW.p)
##### PAUSE here: adjust the network layout on the pop-out window to reach a satisfaction #####
coords <- tk_coords(g_tkid$tkid)
g_tkid <- PWcTalkNW(preNW$PW.pair,preNW$PW.p,layout=coords,pdfW=14,pdfH=10,figname='PWcTalk',asp=0.5)       

```

```{r echo=FALSE, out.width = "90%", fig.align='center', dpi=200, fig.cap="A demonstration pathway crosstalk network"}
result1 <- paste(system.file("extdata", package = "MetaGSCA"), "/PWcTalk.png", sep="")
knitr::include_graphics(result1)
```

<br>




# Citing MetaGSCA
If you use MetaGSCA in your publication, please cite: XXXXXX

















